### 2017-06-01 pkg/model4you/inst/JORS/man_model4you.Rnw
second example, references, size in tikz, ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1425 255f4249-2263-4417-b1ea-8498779e27a7

%% packages
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning}

%% commands

@@

<<setup, echo=FALSE>>=
library("knitr")
knitr::opts_chunk$set(cache = TRUE, message = FALSE)

@@

If you do not believe that all individuals have the same intercept and treatment
effect the R package \emph{model4you} gives you two options to loosen up this 
strict model: 
\paragraph{(1) Model-based trees} identify subgroups where within the subgroups the model 
parameters are similar and between groups the model parameters are different. 
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@

\paragraph{(2) Personalised models} use model-based random forests to estimate similarity
of individuals in terms of model parameters. For each individual a personalised model can
be estimated based on a weighted set of the original data, where the similarity measure 
corresponds to the weight.
<<eval=FALSE>>=
pers_models <- pmodel(base_model)
@

@@

<<GBSG2_prep>>=
library("model4you")
library("survival")
data("GBSG2", package = "TH.data")

set.seed(2017)
@

@@

<<GBSG2_forest>>=
frst <- pmforest(object = bmod, ntree = 100)
@

@@

\section*{References}

\textcolor{blue}{Please enter references in the Harvard style and include a DOI where available, citing them in the text with a number in square brackets, e.g. \\ }

\textcolor{blue}{[1] Piwowar, H A 2011 Who Shares? Who Doesn't? Factors Associated with Openly Archiving Raw Research Data. PLoS ONE 6(7): e18657. DOI: \\ http://dx.doi.org/10.1371/journal.pone.0018657.}

>>>>

%% packages
\usepackage{natbib}

\usepackage{subfig}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning}

@@

<<setup, echo=FALSE>>=
library("knitr")
knitr::opts_chunk$set(cache = TRUE, message = FALSE, fig.align = "center")

@@

If you do not believe that all individuals have the same intercept and treatment
effect the R package \emph{model4you} gives you two options to loosen up this 
strict model: 
\paragraph{(1) Model-based trees} identify subgroups where within the subgroups the model 
parameters are similar and between groups the model parameters are different. This is 
achieved by finding instabilities in the model parameters with respect to a 
variable (characteristicum) and recursively partitioning the data into groups. If, for
example the algorithm finds that men and women have differing treatment effects, the
data is partitioned into two subgroups. Details on model-based trees in general can 
be found in \citep{zeileis_model-based_2008} and for the special use case in clinical 
trial in \citep{seibold_model-based_2016}.

Just a single line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@

\paragraph{(2) Personalised models} use model-based random forests to estimate similarity
of individuals in terms of model parameters. For each individual a personalised model can
be estimated based on a weighted set of the original data, where the similarity measure 
corresponds to the weight. Details on the personalised models can be found in 
\citep{seibold_individual_2017}.

In R, to get from the initial base model to personalised models for all observations
in the training data, type:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@

@@

In the following we will present two example applications for model-based trees
and personalised models. For this we need to load the package and -- to ensure
reproducibility -- set a random seed. Also for plots we will need the \emph{ggplot2}
package and we use a black and white theme.
<<prep>>=
library("model4you")
set.seed(2017)

library("ggplot2")
theme_set(theme_classic())
@

\paragraph{Mathematics exam analysis}
<<math_data>>=
data("MathExam14W", package = "psychotools")

## scale points achieved to [0, 100] percent
MathExam14W$tests <- 100 * MathExam14W$tests/26
MathExam14W$pcorrect <- 100 * MathExam14W$nsolved/13

## select variables to be used
MathExam <- MathExam14W[ , c("pcorrect", "group", "tests", "study",
                             "attempt", "semester", "gender")]
@

<<math_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
## compute base model
bmod_math <- lm(pcorrect ~ group, data = MathExam)
lmplot(bmod_math, densest = TRUE)
@

<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for GBSG2 data.">>=  
## compute tree
(tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2)))
plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = lmplot, 
                                               confint = TRUE))
@

<<math_methods>>=  
## predict
newdat <- MathExam[1:5, ]

# terminal nodes
(nodes <- predict(tr_math, type = "node", newdata = newdat))

# response
(pr <- predict(tr_math, type = "pass", newdata = newdat))

# response including confidence intervals, see ?predict.lm
(pr1 <- predict(tr_math, type = "pass", newdata = newdat,
                predict_args = list(interval = "confidence")))
@

<<math_pmodl>>=
forest_math <- pmforest(bmod_math)
pmod_math <- pmodel(forest_math)
head(pmod_math)
@

<<math_dp1, fig.width=4, fig.height=4, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
dpdat_math <- cbind(pmod_math, MathExam)

ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
  geom_point(alpha = 0.2, size = 1) + 
  geom_smooth(fill = NA, method = "loess") +
  theme(legend.position = "none")

library("ggbeeswarm")
ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
  geom_quasirandom(alpha = 0.5, size = 1) +
  theme(legend.position = "none")
@

<<math_dp2, fig.width=4, fig.height=4, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=
ggplot(dpdat_math, aes(x = attempt, y = group2, color = attempt)) +
  geom_quasirandom(alpha = 0.5) +
  theme(legend.position = "none")

ggplot(dpdat_math, aes(x = gender, y = group2, color = gender)) +
  geom_quasirandom(alpha = 0.5) +
  theme(legend.position = "none")
@

\paragraph{Analysis of the German Breast Cancer Study Group 2}

<<GBSG2_prep>>=
library("survival")
data("GBSG2", package = "TH.data")
@

@@

<<GBSG2_forest>>=
frst <- pmforest(bmod)
@

@@

% \section*{References}
\bibliographystyle{plain}
\bibliography{ref_model4you}

% \textcolor{blue}{Please enter references in the Harvard style and include a DOI where available, citing them in the text with a number in square brackets, e.g. \\ }
% 
% \textcolor{blue}{[1] Piwowar, H A 2011 Who Shares? Who Doesn't? Factors Associated with Openly Archiving Raw Research Data. PLoS ONE 6(7): e18657. DOI: \\ http://dx.doi.org/10.1371/journal.pone.0018657.}
